---
title: "WGCNA - RNASeq"
output: html_document
date: "2023-08-10"
---



```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r Helper Functions, echo = FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}
```


```{r load.packages, echo=FALSE, message=FALSE, results='hide'}

# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")

# if(!require("limma")) BiocManager::install("limma"); library("limma")
library(edgeR)
if(!require("sva")) BiocManager::install("sva"); library("sva")
if(!require("limma")) BiocManager::install("limma"); library("limma")
library("factoextra")
library(WGCNA)
library("data.table")
library(flashClust)
library(DESeq2)
library(tidyverse)
```

# Download dataset 
```{r}
# https://zenodo.org/record/4715907 
expr_dir = "/pastel/Github_scripts/Intro_Systems_Biology/datasets/"

# Metadata 
metadata = as.data.frame(data.table::fread(paste0(expr_dir, "Thoracic_Spinal_Cord/Thoracic_Spinal_Cord_metadata.tsv"), header = T, stringsAsFactors = F, check.names = F))
rownames(metadata) = metadata$rna_id

# Counts expression data 
data = data.table::fread(paste0(expr_dir, "Thoracic_Spinal_Cord/Thoracic_Spinal_Cord_gene_counts.tsv"), header = T, stringsAsFactors = F, check.names = F)
dim(data)
data[1:5, 1:5]

data_tmp = as.data.frame(data)
data_tmp$gene_name = NULL
rownames(data_tmp) = data$ensembl_id
data_tmp$ensembl_id = NULL
data_tmp = data_tmp[, metadata$rna_id] # order the expression table 
identical(colnames(data_tmp), metadata$rna_id) # must be TRUE 
data_tmp[1:3,1:3]
x <- DGEList(counts=as.matrix(data_tmp), samples=metadata)
cpm = cpm(x)
cpm[1:3,1:3]

keep.exp = rowSums(cpm > 1) >= 0.3*ncol(x) #Filtering
x = x[keep.exp,]
dim(x$counts)
counts_filt = as.data.frame(x$counts)
counts_filt[1:3, 1:3]

```

# PCA before adjustment
Counts matrix 
```{r pca, echo=TRUE}
metadata$sex = as.factor(metadata$sex)
metadata$disease = as.factor(metadata$disease)
metadata$library_prep = as.factor(metadata$library_prep)
metadata$site_id = as.factor(metadata$site_id)

res.pca = prcomp(t(counts_filt))
fviz_pca_ind(res.pca, 
             habillage = metadata$disease)
```

## Data normalization
```{r norm, message=FALSE}
dds <- DESeqDataSetFromMatrix(countData = round(counts_filt),
                             colData = metadata,
                             design = ~ disease)

#Using the function of voom normalization
gExpr <- DGEList(counts=assay(dds))
gExpr <- calcNormFactors(gExpr)
vobjGenes <- voom(gExpr, model.matrix( ~ disease, metadata) )
gene_counts_voom = vobjGenes$E # ate aqui
```

# PCA after adjustment
Matrix normalized and adjusted 
```{r pca2}
covariates = c("rin", "pct_mrna_bases", "median_3prime_bias", "median_5prime_bias", "pct_chimeras", "gPC1","gPC2","gPC3","gPC4", "gPC5", "pct_ribosomal_bases", "pct_intergenic_bases", "estimated_library_size" )

residuals <- removeBatchEffect(x = gene_counts_voom, 
                                   batch = metadata$site_id, 
                                   batch2 = metadata$sex, 
                                #   design = model.matrix(~ disease, data = metadata), #force to not regress disease 
                                   covariates = as.matrix(metadata[, covariates])) 

dim(residuals)
res.pca2 = prcomp(t(residuals))
fviz_pca_ind(res.pca2, 
             habillage = metadata$disease)

```

# Clustering: Euclidean
```{r cluster.samples, fig.width=15, fig.height=15}
#Euclidean distance with the transposed matrix
sampleTree = hclust(dist(t(residuals)), method = "average")

plot(sampleTree, main = "Sample clustering to detecting outliers after correction", sub = "Function: hclust, Method: average from Euclidean distance", xlab = "samples", cex.lab = 1.5, cex.axis = 1.5, cex.main = 1.5, cex.sub = 1.5)
```

# Power threshold
```{r soft.threshold, echo=TRUE, fig.width=4, fig.height=8}
# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to=20, by=2))

# Call the network topology analysis function
sft = pickSoftThreshold(t(residuals), powerVector = powers, verbose = 5)

# Scale-free topology fit index as a function of the soft-thresholding power
cex1 = 0.9

plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
main = paste("Scale independence"))
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
labels=powers,cex=cex1,col="red")
# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red")

```

# Connectivity
```{r connectivity, echo=TRUE, fig.width=4, fig.height=5}
# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
```

# TOM adjacency
```{r TOM.adjacency, echo=TRUE, fig.width=8, fig.height=6}
softPower = 7
adjacency = adjacency(t(residuals), power = softPower, type = "signed") 
TOM = TOMsimilarity(adjacency, TOMType="signed")
dissTOM = 1-TOM
geneTree = flashClust(as.dist(dissTOM), method="average")

plot(geneTree, xlab="", sub="", main= "Gene Clustering on TOM-based dissimilarity", labels= FALSE, hang=0.04)

```

# Net parameters
```{r module.size, echo=TRUE, results='asis'}
# "We like large modules" - Tutorial, so we set the minimum module size relatively high:
minModuleSize = 30
# Module identification using dynamic tree cut:
dynamicMods = cutreeDynamic(dendro = geneTree, 
                            distM = dissTOM,
                            deepSplit = 1, 
                            pamRespectsDendro = FALSE,
                            minClusterSize = minModuleSize);

clusters_size = data.matrix(table(dynamicMods))
clusters_size = cbind(rownames(clusters_size),clusters_size)
colnames(clusters_size) = c("cluster","size")

# Convert numeric lables into colors
dynamicColors = labels2colors(dynamicMods)
clusters_colors = data.matrix(table(dynamicColors))
clusters_colors = cbind(rownames(clusters_colors), clusters_colors)
colnames(clusters_colors) = c("color","size")

mytable = table(dynamicColors)
# as.data.frame(t(as.matrix(unclass(mytable))))
```

# Dendrogram colors
```{r dendrogram_net.colors, echo=TRUE, fig.width=10, fig.height=6}
# Plot the dendrogram and colors underneath
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",
dendroLabels = FALSE, hang = 0.03,
addGuide = TRUE, guideHang = 0.05,
main = "Gene dendrogram and module colors")

```

# Number of modules before merge
```{r num_modules, echo=TRUE, results='asis'}
length(unique(dynamicColors))
```

# Clustering eigengenes 
```{r clustering.eigengenes, echo=TRUE, fig.width=12, fig.height=8}
# Calculate eigengenes
resid_expr_t = t(residuals)
MEList = moduleEigengenes(resid_expr_t, colors = dynamicColors)
MEs = MEList$eigengenes
# Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs)
# Cluster module eigengenes
METree = hclust(as.dist(MEDiss), method = "average")

plot(METree, main = "Clustering of module eigengenes",
xlab = "", sub = "")
#We choose a height cut of 0.25, corresponding to correlation of 0.75, to merge
MEDissThres = 0.25
# Plot the cut line into the dendrogram
abline(h=MEDissThres, col = "red")

```

# Merge modules 
```{r merge.modules, echo=TRUE, fig.width=10, fig.height=6}
# Call an automatic merging function
merge = mergeCloseModules(resid_expr_t, dynamicColors, cutHeight = MEDissThres, verbose = 3)
# The merged module colors
mergedColors = merge$colors;
# Eigengenes of the new merged modules:
mergedMEs = merge$newMEs
#png(paste0(expr_dir, "/MyND/Merge_modules.png"), width = 16, height = 8, res = 300, units = "in")
plotDendroAndColors(geneTree, cbind(dynamicColors, mergedColors),
c("Dynamic Tree Cut", "Merged dynamic"),
dendroLabels = FALSE, hang = 0.03,
addGuide = TRUE, guideHang = 0.05)
```

## Merged network
```{r use.merged.network, echo=TRUE, fig.width=10, fig.height=8}
#### For use of the new MERGED DATA!!! 
# Rename to moduleColors
moduleColors = mergedColors
# Construct numerical labels corresponding to the colors
colorOrder = c("grey", standardColors(50));
moduleLabels = match(moduleColors, colorOrder)-1;
MEs = mergedMEs;

```

# Number of modules after merge
```{r num_modules_merg, echo=TRUE, results='asis'}
length(unique(moduleColors))
```

# Trait associations
```{r module_trait.associations_all, echo=TRUE, fig.width=8, fig.height=14 }
metadata_selected2 = metadata[,c("sex", "site_id", "rin", "age_rounded", "disease_duration")]
metadata_selected2$rin = as.numeric(metadata_selected2$rin)
metadata_selected2$site_id = as.numeric(metadata_selected2$site_id)
metadata_selected2$sex = as.numeric(metadata_selected2$sex)
metadata_selected2$age_rounded = as.numeric(metadata_selected2$age_rounded)

nGenes = ncol(resid_expr_t)
nSamples = nrow(resid_expr_t)

# Recalculate MEs with color labels
MEs0 = moduleEigengenes(resid_expr_t, moduleColors)$eigengenes
MEs = orderMEs(MEs0) #Reorder eigenvectors such that similar ones measured by correlation are next to each other
moduleTraitCor = WGCNA::cor(MEs, metadata_selected2, use = "p")
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples)

# Will display correlations and their p-values
textMatrix = paste(signif(moduleTraitCor, 2), "\n(",
signif(moduleTraitPvalue, 1), ")", sep = "")
dim(textMatrix) = dim(moduleTraitCor)
par(mar = c(6, 20, 3, 3));
# Display the correlation values within a heatmap plot
labeledHeatmap(Matrix = moduleTraitCor,
xLabels = names(metadata_selected2), 
yLabels = names(MEs),
ySymbols = names(MEs),
colorLabels = FALSE,
colors = greenWhiteRed(50),
textMatrix = textMatrix,
setStdMargins = FALSE,
cex.text = 0.5,
zlim = c(-1,1),
main = paste("Module-trait relationships"))

gene_ids = colnames(resid_expr_t)
gene_ids = colnames(resid_expr_t)
# save(moduleColors, file = paste0(work_plots, "MyND/moduleColors.Rdata"))
gene_modules = cbind(gene_ids, moduleColors)
gene_modules = as.data.frame(gene_modules)
```

# Gene assignment 
```{r symbol}
## Get conversion table for Gencode 30
gencode_30 = read.table(paste0(expr_dir, "ens.geneid.gencode.v30"))
colnames(gencode_30) = c("ensembl","symbol")
gencode_30$ensembl = gsub("(.*)\\.(.*)","\\1",gencode_30$ensembl)
gene_modules_symbol = merge(gene_modules, gencode_30, by.x = "gene_ids", by.y = "ensembl")

# write.table(gene_modules_symbol, file = paste0(expr_dir, "MyND/geneBymodule.txt"), quote = F, row.names = F, sep = "\t")
createDT( gene_modules_symbol)
```

# Module size 
```{r mod_size}
mytable2 = table(gene_modules_symbol$moduleColors)
createDT(as.data.frame(t(as.matrix(unclass(mytable2)))))
```

And now what? 


